Running an open experiment: transparency and reproducibility in soil and ecosystem

### 2016-06-29 manuscript_openexperiment/Open_experiment.Rmd
Incorporated VLB comments and resubmitted

Vines et al. {, 2014 #4130} published a shocking finding, based on a survey of 516 biology articles from 2 to 22 years old: the odds of a data set being available fell by 17% each year, and the chances that the contact author's email address still worked declined by 7% per year. Similarly, Reichman et al. {, 2011 #2862} estimated that less than 1% of ecological data collected is made available after publication, and noted, as an example, that much current and historical data relevant to the 2010 *Deepwater Horizon* oil spill are already inaccessible or lost {Reichman, 2011 #2862}. In global change ecology, Wolkovich et al. {, 2012 #3303} reported that they were able to acquire only ~10% of other researchers' raw data sets in preparation for a meta-analysis {Wolkovich, 2012 #3331}.

@@

There were thus significant advantages to implementing and using an automated analytical pipeline in an open repository. It is worth considering in depth, however, the costs of such an approach, the investments required, and the barriers we encountered (or that many other scientists would).

* More upfront time is required in an open repository system, as the basic building blocks are assembled into an automated pipeline that runs from raw data to diagnostics to final products. While frustrating, this is generally a good thing however: automation makes later analytical stages much faster (see above), while openness provides a strong incentive to write clean, clear code from the beginning, for example, as opposed to deferring documentation.
* To our knowledge, there is no 'template' for an experimental repository such as the one used here. This results in unnecessary time re-inventing wheels, in particular with respect to repository design decisions.
* Effective data management takes some programming skills. We believe that basic programming, like basic statistics, should be a part of almost every scientist's training, and do not view this as a huge problem going forward. For scientists used to spreadsheet-oriented data processing, however, it requires new skills and practices.
* Version control (here, the Git software) is critical for tracking provenance, ensuring robustness, and efficiently sharing changes. It introduces significant complexity, however, particularly for any use beyond the most simple operations. We believe version control software remains too difficult for most scientists to use effectively[^9].
* Git and GitHub are fundamentally designed for working with code, not data or long documents. This situation continues to improve on a technical level[^10], but some operations (e.g., storing very large files; tracking small changes to columnar data; commenting on words or phrases) remain awkward or even impossible.

@@

A number of caveats are also in order. The model we present here--a public repository updated throughout the experiment, analysis, and publication process--is far from perfect. For example, there are a number of "best practices" of scientific computing {Wilson, 2014 #4337} that we did not employ, in particular with respect to automation. In addition, this approach may not be appropriate for, nor applicable to, all study types, even in our professional fields of soil, ecosystem, and global change science; it certainly is not appropriate for sensitive human-subject data.

More generally, by itself transparency does not guarantee repeatability and reproducibility {Easterbrook, 2014 #4137}, and it may raise genuine concerns about, for example, protecting scientists from harassment {Lewandowsky, 2016 #4342}. Nor will open science by itself fundamentally enhance public trust in science, as there are still many social and political challenges to overcome {Grand, 2012 #4370}.

Scientists also frequently cite other concerns: about data sharing, being "scooped", not receiving sufficient credit, and time constraints {Wolkovich, 2012 #3303}. The benefits are not always clear to researchers who might otherwise be open to following open-data practices {Molloy, 2011 #4366}{Hampton, 2015 #4420}. We cannot easily dismiss all of these concerns: for example, does our community adequately credit researchers who contribute to global databases that subsequently produce high-impact papers? Such meta-analyses rely on the collection of primary data, and it is critical that field and experimental researchers' efforts are adequately valued and cited {Kueffer, 2011 #4335}.

Finally, we recognize that, in general, our professional and career incentives do not yet align well with "an open research culture" {Nosek, 2015 #4343}. That is, scientists do not receive 'proper' credit for datasets relative to publications, even when the former may be much more broadly valuable, in almost every area: hiring, promotion and tenure decisions, impact metrics, etc. The advent of 'data descriptor' articles, in journals such as *Scientific Data*, improves but does not solve this problem.

>>>>

Vines et al. {, 2014 #4130} published a shocking finding, based on a survey of 516 biology articles from 2 to 22 years old: the odds of a data set being available post-publication fell by 17% each year, and the chances that the contact author's email address still worked declined by 7% per year. Similarly, Reichman et al. {, 2011 #2862} estimated that less than 1% of ecological data collected is made available after publication, and noted, as an example, that much current and historical data relevant to the 2010 *Deepwater Horizon* oil spill are already inaccessible or lost {Reichman, 2011 #2862}. In global change ecology, Wolkovich et al. {, 2012 #3303} reported that they were able to acquire only ~10% of other researchers' raw data sets in preparation for a meta-analysis {Wolkovich, 2012 #3331}.

@@

There were thus significant advantages to implementing and using an automated analytical pipeline in an open repository. It is worth considering in depth, however, the costs of such an approach, the investments required, and the barriers encountered.

* More upfront time is required in an open repository system, as the basic building blocks are assembled into an automated pipeline that runs from raw data to diagnostics to final products. While frustrating, this is generally good: automation makes later analytical stages much faster (see above), while openness provides a strong incentive to write clean, clear code from the beginning, for example, as opposed to deferring documentation.
* To our knowledge, there is no 'template' for an experimental repository such as the one used here. This results in unnecessary time spent "re-inventing wheels", in particular with respect to repository design decisions. Clear, flexible, and powerful templates will help scientists take advantage of repository-based and open science approaches.
* Effective data management in a repository system takes some programming skills. Basic programming is becoming an increasingly important part of most scientists' training, just as basic statistics has long been a critical skill. For scientists accustomed to spreadsheet-based data processing, investment in new skills and practices, or new partnerships with skilled programmers, will facilitate more open analyses.
* Version control (here, the Git software) is critical for tracking provenance, ensuring robustness, and efficiently sharing changes. It introduces significant complexity, however, particularly for any use beyond the most simple operations. In our experience, version control software remains too difficult for most scientists to use effectively[^9]. We urge software developers to work closely with diverse science communities to develop more user-friendly tools.
* Git and GitHub are fundamentally designed for working with code, not data or long documents. This situation continues to improve on a technical level[^10], but some operations (e.g., storing very large files; tracking small changes to columnar data; commenting on words or phrases) remain awkward or even impossible.

@@

A number of caveats are also in order. The model we present here--a public repository updated throughout the experiment, analysis, and publication process--is far from perfect. For example, there are a number of "best practices" of scientific computing {Wilson, 2014 #4337} that we did not employ, in particular with respect to automation. In addition, this approach may not be appropriate for, nor applicable to, all study types, even in our professional fields of soil, ecosystem, and global change science; it certainly is not appropriate for sensitive data, e.g. human-subject research.

More generally, by itself transparency does not guarantee repeatability and reproducibility {Easterbrook, 2014 #4137}, and it may raise new concerns about, for example, protecting scientists from harassment {Lewandowsky, 2016 #4342}. Nor will open science by itself fundamentally enhance public trust in science, as there are still many social and political challenges to overcome {Grand, 2012 #4370}.

Scientists also frequently cite other concerns: about data sharing, being 'scooped', not receiving sufficient credit, and time constraints {Wolkovich, 2012 #3303}. The benefits are not always clear to researchers who might otherwise be open to following open-data practices {Molloy, 2011 #4366}{Hampton, 2015 #4420}. We cannot easily dismiss all of these concerns: for example, does our community adequately credit researchers who contribute to global databases that subsequently produce high-impact papers? Such meta-analyses rely on the collection of primary data, and it is critical that field and experimental researchers' efforts are adequately valued and cited {Kueffer, 2011 #4335}.

Finally, we recognize that, in general, our professional and career incentives do not yet align well with "an open research culture" {Nosek, 2015 #4343}. That is, scientists do not receive appropriate credit for datasets relative to publications, as hiring, promotion and tenure decisions all tend to reward publications, not datasets. The advent of 'data descriptor' articles, in journals such as *Scientific Data*, improves but does not solve this problem.
